---
title: "Biomarkers of ASD"
subtitle: "If you want a subtitle put it here"
author: "List names here"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. In all, your report can be quite short. When it is complete, render and then push changes to your team repository.

```{r}
# load any other packages and read data here
library(tidyverse)
```

## Abstract

Write a brief one-paragraph abstract that describes the contents of your write-up.

## Dataset

Write a brief data description, including: how data were obtained; sample characteristics; variables measured; and data preprocessing. This can be largely based on the source paper and should not exceed 1-2 paragraphs.


## Summary of published analysis

Summarize the methodology of the paper in 1-3 paragraphs. You need not explain the methods in depth as we did in class; just indicate what methods were used and how they were combined. If possible, include a diagram that depicts the methodological design. (Quarto has support for [GraphViz and Mermaid flowcharts](https://quarto.org/docs/authoring/diagrams.html).) Provide key results: the proteins selected for the classifier and the estimated accuracy.

## Findings

Summarize your findings here. I've included some subheaders in a way that seems natural to me; you can structure this section however you like.

### Impact of preprocessing and outliers

```{r}
library(readr)

load("../data/biomarker-clean.RData")

# clean data without log-transformation
var_names <- read_csv('../data/biomarker-raw.csv', 
                      col_names = F, 
                      n_max = 2, 
                      col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, 
         abbreviation = V2) %>%
  na.omit()

biomarker_raw <- read_csv('../data/biomarker-raw.csv', 
                            skip = 2,
                            col_select = -2L,
                            col_names = c('group', 
                                          'empty',
                                          pull(var_names, abbreviation),
                                          'ados'),
                            na = c('-', '')) %>%
  filter(!is.na(group))

```

```{r}
#investigate the potential reason for log-transformation
# check normality for some protein

par(mfrow = c(1, 2))

qqnorm(main = "Raw Data Q-Q plot for `Mcl-1`", biomarker_raw$`Mcl-1`)
qqline(biomarker_raw$`Mcl-1`, col = "red")
qqnorm(main = "Clean Data Q-Q plot for `Mcl-1`",biomarker_clean$`Mcl-1`)
qqline(biomarker_clean$`Mcl-1`, col = "red")

```

```{r}
par(mfrow = c(1, 2))
qqnorm(main = "Clean Data Q-Q plot for `DERM`",biomarker_raw$DERM)
qqline(biomarker_raw$DERM, col = "red")
qqnorm(main = "Clean Data Q-Q plot for `DERM`",biomarker_clean$DERM)
qqline(biomarker_clean$DERM, col = "red")
```

We compare the data before processing and after processing. Specifically, we compare the raw data set and data after log-transformation and scaling. We randomly selected two sample protein type and examine the normality. It could be seen that `DERM` is normally distributed before pre-porcessing but not centered at 0, while `Mcl-1` is neither normally distributed nor centralized. Therefore, we could conclude that not all protein types are normally distributed in the raw data, so a log-transformation is used. In this way, the data used for further model training could be more reliable.

```{r}

```

### Methodlogical variations

Task 3

### Improved classifier

Task 4
